Unicode Implementation Issues \thanks { This is a working document for the Rexx Language Association (RexxLA). There will be one or more separate requirements documents. It is written in \LaTeX\ \fmtversion\ and was rendered \today\ using \enginedetails. }

### 2023-10-31 Unicode.Implimentation.tex
Add files via upload

\subsection{Tools}
There are tools available that manipilate Rexx source code in various
ways, e.g., code completion, cross referencing, navigation, prettyprinting, syntax highligting.
These tools may not handle new syntax added in support of Unicode. Examples include
\begin{itemize}
\item The IBM Z\textregistered Open Editor \\
A Rexx Language Support plugin for Visual Studio Code
\item The \LaTeX\  package \href{https://www.ctan.org/pkg/listings}{\textbf{listings}}
\item The \LaTeX \ package \href{https://www.ctan.org/pkg/minted}{\textbf{minted}}
\item The Python package Pygments
\end{itemize}

@@

RXU has five types of string literals:
\begin{definition}
\item['string'] The type is determined by OPTIONS.
\item['string'P] Unicode by code point
\item['string'T] Unicode by EGC
\item['space separated tokens'U]
Unicode code points by alias, name or hexadecimal value;
the string is BYTES rather than CODEPOINTS or TEXT.
\item['string'Y] Legacy bYte string
\end{definition}

@@

@@

Case folding of symbols and caseless comparisons will use the \citetitle{UCD}.
\cref{issue:CF,issue:CComp,issue:CConv})
The rules for case folding should take into account mathematical usage.
\begin{itemize}
\item U+1D400 through U+1D7FF are semantically distinct from other letters.
\item Superscripts and subscripts have semantic significance.
\end{itemize}

@@

>>>>

\subsection{Tools}
There are tools available that manipilate Rexx source code in various
ways, e.g., code completion, cross referencing, navigation, prettyprinting, syntax highligting.
These tools may not handle new syntax added in support of Unicode. Examples include
\begin{itemize}
\item The IBM Z\textregistered Open Editor \\
A Rexx Language Support plugin for Visual Studio Code
\item The \LaTeX\  package \href{https://www.ctan.org/pkg/listings}{\textbf{listings}}
\item The \LaTeX \ package \href{https://www.ctan.org/pkg/minted}{\textbf{minted}}
\item The Python package \href{https://pygments.org/}{\textbf{Pygments}}
\end{itemize}

@@

\subsubsection{RXU string types}

RXU has five types of string literals\cite{RXU:NewTypes}:
\begin{definition}
\item['string'] The type is determined by OPTIONS.
\item['string'P] Unicode by code point
\item['string'T] Unicode by EGC
\item['space separated tokens'U]
Unicode code points by alias, name or hexadecimal value;
the string is BYTES rather than CODEPOINTS or TEXT.
\item['string'Y] Legacy bYte string,can only be promoted if valid UTF-8 sequences
\end{definition}

@@

\subsubsection{RXU I/O}

in RXU, stream I/O\cite{RXU:STREAM} defaults to legacy (BYTES).

RXU adds an \textbf{ENCODING} option to \textbf{STREAM(}\textit{streamid},\textbf{OPEN }\textit{options}\textbf{)}. When this option is used the stream is Unicode enabled, and some stream functions are not allowed.
\textbf{ENCODING} must be followed by the name of then encoding, which for Unicode may in turn be followed by \textit{error} and \textit{target} options:

\begin{itemize}
  \item \underline{\textbf{REPLACE}} \\
  Replace invalid characters with the Unicode Replacement Character (U+FFFd).
  \item \textbf{SYNTAX} \\
  Raise SYNTAX for invalid characters.
  \item \textbf{CODEPOINTS} \\
  Units of codepoints.
  \item \underline{\textbf{TEXT}} \\
  Units of extended grapheme clusters.
\end{itemize}

RXU adds  options to  \textbf{STREAM(}\textit{streamid},\textbf{QUERY}\textit{options}\textbf{)}

\begin{itemize}
\item \textbf{Query Encoding Name}
\item \textbf{Query Encoding Target} \\
CODEPOINTS  or TEXT
\item \textbf{Query Encoding Error} \\
REPLACE or SYNTAX
\item \textbf{Query Encoding LastError}
\item \textbf{Query Encoding \\
Everything but LastError}
\end{itemize}

@@

Case folding of symbols and caseless comparisons will use the \citetitle{UCD}ENCODING.
\cref{issue:CF,issue:CComp,issue:CConv})
The rules for case folding should take into account mathematical usage.
\begin{itemize}
\item U+1D400 through U+1D7FF are semantically distinct from other letters.
\item Superscripts and subscripts have semantic significance.
\end{itemize}

@@

\subsubsection{RXU functions and methods}

RXU adds the \textbf{DECODE} and \textbf{ENCODE} methods in the Unicode class. These provide an easy way to implement the proposed \textbf{iconv} function.

RXU adds these\cite{RXU:NewBIFs} additional built-in functions:

\begin{itemize}
\item \textbf{BYTES}
\item \textbf{C2U}
\item \textbf{CODEPOINTS}
\item \textbf{GRAPHEMES}
\item \textbf{N2P}
\item \textbf{P2N}
\item \textbf{STRINGTYPE}
\item \textbf{TEXT}
\item \textbf{UNICODE(}\textit{string},\textit{function}\textbf{)} \\
Performs the named function on \textit{string}:
\begin{description}
\item[isNFD] Returns 1 if string is in normalized form D
\item[toLowercase]
Returns upper case translatio.
\item[toUpercase]
Returns lower case translation
\item[toNFD] Normalizes to form D (decomposed)
\end{description}
\item \textbf{UNICODE(}\textit{codepoint},\textbf{PROPERTY},\textit{property}\textbf{)} \\
Retrieves the value of the named property
\item \textbf{UTF8} \\
Validates a UTF-8 string and optionally converts it to a different Unicode transform.
\end{itemize}
